---
title: "The Impact of Weight on a Sumo Wrestler's Success"
author: "Nazeem Kamel"
date: today
format: html
execute:
  echo: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(broom)
#| label: clean-data
#| include: false

library(tidyverse)

# Download the original data once
dir.create("data", showWarnings = FALSE)

data_file <- "data/sumo_since_1957.csv"

if (!file.exists(data_file)) {
  download.file(
    "https://data.scorenetwork.org/data/sumo_since_1957.csv",
    destfile = data_file,
    mode = "wb"
  )
}

sumo_raw <- read_csv(data_file, show_col_types = FALSE)

# Create a separate working dataset
sumo_clean <- sumo_raw |>
  mutate(
    # Remove unnecessary whitespace in wrestler names
    Rikishi = str_squish(Rikishi),
    
    # Convert birth dates from text into dates
    birth_date = as.Date(`Birth Date`, format = "%d.%m.%Y"),
    
    # Convert height to metres
    height_m = height_cm / 100,
    
    # Calculate BMI only when height and weight are valid
    bmi = if_else(
      is.finite(weight_kg) & weight_kg > 0 &
        is.finite(height_m) & height_m > 0,
      weight_kg / height_m^2,
      NA_real_
    )
  )
```

```{r}
#CREATE TABLE OF PERFORMANCE ROWS THAT RECORD FULL 15 BOUTS 
library(dplyr)

sumo_15 <- sumo_clean |>
  mutate(
    recorded_bouts = wins + losses
  ) |>
  filter(
    recorded_bouts == 15
  ) |>
  mutate(
    win_pct = 100 * wins / recorded_bouts
  )
```

```{r}
#TEST FOR BMI AGAINST INCOMPLETE TOURNEY COUNT
#library(dplyr)
#
#sumo_by_rikishi <- sumo_clean |>
#  group_by(Rikishi) |>
#  summarise(
#    avg_win_pct = mean(win_pct),
#    appearances = n(),
#    .groups = "drop"
#  )
#
#View(sumo_by_rikishi)
```

```{r}
#GROUP BY SUMO WRESTLERS AND INCLUDE BMI STAT
library(dplyr)

sumo_by_rikishi_15 <- sumo_15 |>
  group_by(Rikishi, birth_date = .data[["Birth Date"]]) |>
  summarise(
   avg_bmi = mean(bmi),
    avg_win_pct = mean(win_pct),
    appearances = n(),
    .groups = "drop"
  )
#View(sumo_by_rikishi_15)
```

```{r}
library(dplyr)

sumo_wrestlers <- sumo_15 |>
  group_by(Rikishi, birth_date = .data[['Birth Date']]) |>
  summarise(
    avg_bmi = mean(bmi),
    avg_weight_kg = mean(weight_kg),
    avg_win_pct = mean(win_pct),
    appearances = n(),
    .groups = "drop"
  )
#View(sumo_wrestlers)
```

```{r}
sumo_2plus <- sumo_wrestlers |>
  filter(appearances >= 2)
```

```{r}
# All retained Rikishi groups
bmi_all <- lm(
  avg_win_pct ~ avg_bmi,
  data = sumo_wrestlers
)

weight_all <- lm(
  avg_win_pct ~ avg_weight_kg,
  data = sumo_wrestlers
)

# Groups with at least two qualifying appearances
bmi_2plus <- lm(
  avg_win_pct ~ avg_bmi,
  data = sumo_2plus
)

weight_2plus <- lm(
  avg_win_pct ~ avg_weight_kg,
  data = sumo_2plus
)

# Inspect the restricted-sample results
summary(bmi_2plus)
summary(weight_2plus)

# Confidence intervals for the weight model
confint(weight_2plus)
```

```{r}
#FIT DA MODEL
bmi_model <- lm(
  avg_win_pct ~ avg_bmi,
  data = sumo_by_rikishi_15
)

summary(bmi_model)
```

```{r}
library(ggplot2)

ggplot(sumo_by_rikishi_15,
       aes(x = avg_bmi, y = avg_win_pct)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE) +
  labs(
    title = "Average BMI and Winning Percentage",
    x = "Average BMI (kg/m²)",
    y = "Average winning percentage (%)"
  ) +
  theme_minimal()
```

# Abstract

The term size matters is

# Introduction

Professional sumo wrestling is an incredibly old sport, dating back over a thousand years. In its inception, the concept of weight classes did not exist, and it has stuck to these origins until today. Sumo is one of the few organized combat sports one could observe where there is truely no regulation on the size of the competitor. A wrestler wins by forcing his opponent out of the ring, or forcing him to touch the ground with anything but the soles of his feet. Obviously, this causes there to be some implication that size is an important characteristic. This study asks whether wrestlers with a hgher recorded body mass on average also show a higher average winning percentage across tournaments.

To investigate this question, records from the Makuuchi division are used to calculate a wrestler's average recorded body weight at the time of each of his tournament appearances, as well as the percentage of matches won from said appearances. Linear regression is used to examine the relationship between these two averages. Initially, BMI was pursued as a possible parameter, with weight being an alternative predictor.

# Data

The data was obtained from SCORE's Sumo Wrestler Characteristics dataset. The repository gathers records of wrestler performances from 1957 until 2023. The original dataset contains 15,760 rows and 10 columns, with each row being a recording of a sumo wrestler's participation in a Makuuchi tournament, with relevant parameters being their ring name, height, weight, wins, and losses.

The predictor used is body weight in kilograms, with the response being the average winning percentage. Winning percentage was calculated by dividing the wins over the sum of wins and losses and mulitplying be 100. Initially, BMI was pursued as a potential predictor, but weight was ultimately decided upon. The weight and winning percentages of each sumo wrestler were then averaged across all recorded appearances that could be found under that ring name.

However, only appearances that recorded the full 15 matches being completed were recorded, leaving 14,547 rows total. This excludes tournament records that were ended prematurely, presumably by expulsion, injury, or retirement. This allows the win percentages to all share the same denominator of 15 matches total. Additionally, any appearances with missing values in their fields were not used.

The analysis groups appearances by recorded ring name, resulting in 517 wrestlers. The CSV lacks tournament dates, limiting the ability to account for potentially significant variables such as age at competition, or differences in performances across historical eras, which might favor certain builds due to competition meta or ruleset at the time.

\`\`\`{r}

#\| label: tbl-descriptive

#\| tbl-cap: "Descriptive statistics for groups with at least two qualifying appearances."

knitr::kable(

descriptives, digits = 2,

col.names = c("Variable", "Mean", "SD", "Minimum", "Maximum"),

booktabs = TRUE

)

\`\`\`

# Methods

Tournament appearances were filtered to only include those whos recorded wins and losses added up to 15. For each retained appearance, the winning percentage was calculated as (100 \\times \\text{wins}/(\\text{wins}+\\text{losses})). Records were then grouped by a combination of Rikishi name and birth date, as to seperate wrestlers who shared a ring name, of which there were found to be 27. Average bodyweight and average win percentage were calculated across all complete tournament appearances by a wrestler. The main analysis required at least two complete tournament appearances as to ensure a more consistent average was representative by each participating wrestler.

Linear regression was used to examine the association between average bodyweight and average win percentage, the model was:

\[\
Y_i=\\beta_0+\\beta_1X_i+\\varepsilon_i,\
\]

where (Y_i) is the average win percentage for group (i), (X_i) is the avg bodyweight in kilograms, (\\beta_0) is the intercept, and (\\beta_1) is the slope. The error, (\\varepsilon_i), represents the variation not explained by the relationship of the two parameters. The slope describes the expected difference in average win percentage be point of percent for each kilogram of average bodyweight across tournaments.

BMI was initially explored as a measure of body size. It was calculated for each appearance as weight in kilograms divided by squared height in meters, then averaged within groups. Separate regressions were done using avg BMI and avg weight. Both models were also fitted with groups including single tournament appearances included to see how their inclusion would affect results.

Using name and birth date separates shared names but does not resolve name changes within a wrestler’s career. Some wrestlers may end up contributing more than one group, limiting independence in rows. The analysis also does not account for age, historical era, rank, or opponent/tournament strength. The results describe associations between group averages and don't establish that gaining weight improves an individual wrestler’s performance.

# Results

# Discussion

# References
